---
title: "Flights Do Not Have To Be Frustrating Anymore"
output: slidy_presentation
date: "2024-11-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Air travel saw significant growth in the mid-20th century, and the **AirPassengers** dataset offers a glimpse into this trend. This dataset contains monthly passenger counts for an international airline from 1949 to 1960 and finally all the way up to 2013. This project will explore the dataset to discover long-term trends, seasonal variations, and patterns in airline passenger numbers.

---

# Problem
This project addresses the challemnges of air travel. Air travel is a very important component of global connectivity and if the passengers are aware of the challenges that come with it, they can streamline their travel experience. Some of the challenges include delays, overbooking, missing baggages and etc.
We will be manipulating nycflights13 and airpassengers database to tackle our challenge and find out what flights we can pick to avoid running into delays.

---

# Approach
We are using **nycflights13** and **AirPassengers** database, additionally: data visualization tools, such as ggplot2, are deployed to extract seasonal trends, visualize departure delays, and identify patterns in airline operations.
---


# Dataset Overview

The **AirPassengers** dataset is a built-in time series dataset in R. It contains monthly data spanning 12 years, with variables representing the year, month, and passenger counts.

```{r}
data("AirPassengers")

# Convert it into a dataframe
airpassengers_df <- data.frame(
  Year = rep(1949:1960, each = 12),
  Month = rep(month.name, times = 12),
  Passengers = as.numeric(AirPassengers)
)

head(airpassengers_df)


```


```{r}
summary(airpassengers_df)

```

# Monthly Airline Passengers

Next chunk of code creates a continuous time variable by adding the fractional part of each month to the year, adjusted to center each month around its midpoint. This makes the time suitable for smooth plotting and trend analysis.

```{r}
library(ggplot2)
ggplot(airpassengers_df, aes(x = Year + (1:12)/12 - 1/24, y = Passengers)) +
  geom_line(color = "steelblue", line_width = 1) +
  labs(
    title = "Monthly Airline Passengers (1949-1960)",
    x = "Year",
    y = "Passengers"
  )

```


# Seasonal Trends

```{r}
ggplot(airpassengers_df, aes(x = Month, y = Passengers)) +
  geom_boxplot(fill = "lightblue", color = "darkblue") +
  labs(
    title = "Seasonal Trends in Air Travel",
    x = "Month",
    y = "Passengers"
  )

```

# Flight Delays In The new Century (2013)



```{r}
library(nycflights13)
library(dplyr)

flights_with_airlines <- flights %>%
  filter(!is.na(dep_delay)) %>%  
  left_join(airlines, by = "carrier")

ggplot(flights_with_airlines, aes(x = reorder(name, dep_delay), y = dep_delay)) +
  geom_boxplot(outlier.color = "red", fill = "lightblue") +
  coord_flip() +
  labs(
    title = "Distribution of Departure Delays by Airline",
    x = "Airline",
    y = "Departure Delay (minutes)"
  )

```

# What can we deduce from this?

The box plot reveals that some airlines consistently have shorter delays, making them a better choice for timely travel when visiting relatives. To avoid disruptions, consider flying earlier in the day, as delays tend to accumulate later due to cascading scheduling issues. Additionally, avoid airlines with frequent extreme delays (outliers) for smoother travel experiences.


```{r}
library(ggplot2)

average_delay <- flights_with_airlines %>%
  group_by(name) %>%
  summarise(avg_delay = mean(dep_delay, na.rm = TRUE)) %>%
  arrange(desc(avg_delay))

ggplot(average_delay, aes(x = reorder(name, avg_delay), y = avg_delay)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Departure Delay by Airline",
    x = "Airline",
    y = "Average Departure Delay (minutes)"
  ) 
```

Summary: Challenges and Observations in Air Travel
Air travel has many issues, such as flight delays, cancellations, overbooking, and problems with baggage. Our study of the AirPassengers data shows a longterm increase in airline passengers, with seasonal highs and lows that match general travel trends. By examining flight delay data from 2013, we found out how delays differ between airlines, identifying which ones have more frequent or severe delays.

The box plots in our analysis show significant differences in departure delays among airlines. Some airlines consistently have delays, while others have better on-time performance. This information can help travelers choose airlines more wisely, possibly improving their travel experience by avoiding carriers with frequent delays or extreme delays.

Summary: Impacts and Restrictions
The results of this study have significant effects for both airlines and travelers. Airlines can use this information to enhance their operations, make their schedules more efficient, and cut down on delays. For travelers, knowing about delay patterns can help them make smarter choices, like picking airlines with fewer delays or booking earlier flights to avoid long waits. This knowledge could also lead to better policies for making air travel more efficient and dependable.

But it's important to remember the limitations of this analysis. The AirPassengers dataset only includes monthly airline passenger numbers from 1949 to 1960, which might not reflect current air travel trends. Also, the flight delay data from 2013 doesn't consider factors like weather, technical problems, or overall airport delays, which could affect the results. Moreover, the dataset does not include information about locations, so we can't see how different airports or areas affect flight delays. This limits our analysis to just airline performance without looking at the bigger picture of how operations work in different places.
